### 2024-11-05 paper.qmd
update on paper methods for two-stage modeling

To enhance the exploration of parameter space, each subject´s RRi data was standardize against his own mean and standard deviation, improving convergence and an efficient exploration of the posterior distribution in this high dimensional parameter space.

The priors were chosen based on physiological constraints and graphical visualization of standardized RRi data. These priors are detailed in @eq-priors, ensuring in this way, identifiability of model parameters by constraining the parameter space to physiologically plausible values and ultimately, improving model convergence.

$$
\begin{aligned} 
\alpha &\sim \mathcal{N}(1, 0.5) \\
\beta &\sim \mathcal{N}(-2.5, 0.5) \text{ with } \beta \leq 0 \\
c &\sim \mathcal{N}(0.8, 0.2) \text{ with } c \geq 0 \\
\lambda &\sim \mathcal{N}(-2, 0.5) \text{ with } \lambda \leq 0 \\
\phi &\sim \mathcal{N}(-2, 0.5) \text{ with } \phi \leq 0 \\
\tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
\delta &\sim \mathcal{N}(5, 0.5) \text{ with } \delta \geq 0
\end{aligned}
$$ {#eq-priors}

@@

[**Figure 1**]{#fig-prior-sim}. Predicted RRi response to exercise, illustrating uncertainty in model parameters from the prior distribution. Shaded areas represent 95%, 80%, and 60% quantile CI.

The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 25,000 post-warmup samples. The model was fitted for each subject´s standardized RRi data, allowing the estimation of the model parameters associated with individual RRi dynamics for each subject. Afterwards, we transformed back to the RRi scale the $\alpha$ and $\beta$ parameters, ensuring a physiologically meaningful interpretation of model estimation.

@@

[**Figure 5**]{#fig-2d-kernel-density}. 2D kernel density of RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol. Darker colors indicate greater probability density. The contrary can be said about lighter colors.

@@

In this study, we present a non-linear model designed to capture R-R interval (RRi) fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics. By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for both clinical and athletic applications. This model marks a step forward in computational physiology, bridging the gap between theoretical frameworks and practical monitoring applications.

>>>>

The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 25,000 post-warmup samples.

To enhance the exploration of parameter space we performed a two-stage analysis: individual-level and group-level estimates.

### Individual-level analysis

Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardize against his own mean $\bar{\text{RRi}_i}$ and standard deviation $\sigma_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution. The standardized RRi data $y_{i,t}$ for each time point $t$ was computed as:

$$
y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{\sigma_{\text{RRi}_i}}
$$
This standardization allowed the model to focus on relative changes in RRi dynamics, independent of individual baseline differences.

The model for each subject $i$ was then specified in terms of standardized RRi data $y_{i,t}$:

$$
y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-\beta_i \cdot c}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
$$
where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.

Afterwards, we transformed the estimated $\alpha$ and $\beta$ parameters back to the original RRi scale, ensuring a physiologically meaningful interpretation. The transformation for each subject $i$ is given by:

$$
\begin{aligned}
\alpha_i^{\text{RRi}} &= \alpha_i \cdot \sigma_{\text{RRi}_i} + \bar{\text{RRi}}_i \\
\beta_i^{\text{RRi}} &= \beta_i \cdot \sigma_{\text{RRi}_i}
\end{aligned}
$$

Priors were chosen based on physiological constraints and graphical visualization of standardized RRi data, ensuring identifiability of model parameters by constraining the parameter space to plausible values, which improves model convergence. The prior distributions are defined as follows:

$$
\begin{aligned} 
\alpha &\sim \mathcal{N}(1, 0.5) \\
\beta &\sim \mathcal{N}(-2.5, 0.5) \text{ with } \beta \leq 0 \\
c &\sim \mathcal{N}(0.8, 0.2) \text{ with } c \geq 0 \\
\lambda &\sim \mathcal{N}(-2, 0.5) \text{ with } \lambda \leq 0 \\
\phi &\sim \mathcal{N}(-2, 0.5) \text{ with } \phi \leq 0 \\
\tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
\delta &\sim \mathcal{N}(5, 0.5) \text{ with } \delta \geq 0
\end{aligned}
$$

@@

[**Figure 1**]{#fig-prior-sim}. Prior predicted RRi response to exercise, illustrating uncertainty in model parameters from the prior distribution. Shaded areas represent 95%, 80%, and 60% quantile CI.

### Group-level analysis

After obtaining the posterior distribution for each subject's parameters, the mean and standard deviation for each parameter were calculated. These estimates were then used as input data to create a multivariate hierarchical model, capturing variability at both the subject and group levels. The model is described as follows:

For each subject $i$, the parameter vector $\theta_i = (\alpha_i, \beta_i, c_i, \lambda_i, \phi_i, \tau_i, \delta_i)^\top$ is modeled as:

$$
\theta_{i,j} \sim \mathcal{N}(\mu_{j} + b_{i,j}, \sigma_j^2)
$$

where $\mu_{j}$ is the group-level mean for parameter $j$, $b_{i,j}$ represents the subject-level random effect for subject $i$ on parameter $j$, and $\sigma_j^2$ is the residual variance for parameter $j$.

The subject-level effects $b_{i,j}$ are assumed to follow a multivariate normal distribution:

$$
\mathbf{b}_i = (b_{i, \alpha}, b_{i, \beta}, b_{i, c}, b_{i, \lambda}, b_{i, \phi}, b_{i, \tau}, b_{i, \delta})^\top \sim \mathcal{N}(\mathbf{0}, \Sigma_b)
$$

where $\Sigma_b$ is the covariance matrix that captures correlations between the random effects for different parameters across subjects.

This hierarchical structure enables us to capture individual variability through subject-level random effects while also estimating group-level effects across all parameters, thus providing estimates into both subject-level and population-level model parameters.

@@

[**Figure 5**]{#fig-2d-kernel-density}. Upper panel shows the mean and standard deviation (SD) from each of the subject's RRi recordings, used for the standardization process. Lower panel indicate the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol. Darker colors indicate greater probability density. The contrary can be said about lighter colors.

@@

In this study, we present a non-linear model designed to RRi fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics. By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for both clinical and athletic applications. This model marks a step forward in computational physiology, bridging the gap between theoretical frameworks and practical monitoring applications.
